---
title: "Evaluating Biomarkers: Connecting the Theory to Practice"
author: "Michael Sachs"
date: October 30, 2014
output: 
  ioslides_presentation:
    self_contained: false
    widescreen: true
    fig_width: 8
    fig_height: 4
    logo: US-NIH-NCI-Logo.svg
    css: styles.css
---

# Introduction

## Evaluating Biomarkers: Practice

```{r libs, include = FALSE}
library(ggplot2)
library(plyr)
library(ggthemes)
library(survival)
library(survivalROC)
library(plotROC)

ggkm <- function(sfit, xlabs = "Time since marker measured", ylabs = "survival probability", 
                 ystratalabs = NULL, ystrataname = NULL, timeby = 100, 
                 main = "Kaplan-Meier Plot", pval = TRUE, ...) {

if(is.null(ystratalabs)) {
   ystratalabs <- as.character(levels(summary(sfit)$strata))
}
m <- max(nchar(ystratalabs))
if(is.null(ystrataname)) ystrataname <- "Strata"
times <- seq(0, max(sfit$time), by = timeby)
.df <- data.frame(time = sfit$time, n.risk = sfit$n.risk,
    n.event = sfit$n.event, surv = sfit$surv, strata = summary(sfit, censored = T)$strata,
    upper = sfit$upper, lower = sfit$lower)
levels(.df$strata) <- ystratalabs
zeros <- data.frame(time = 0, surv = 1, strata = factor(ystratalabs, levels=levels(.df$strata)),
    upper = 1, lower = 1)
.df <- rbind.fill(zeros, .df)
d <- length(levels(.df$strata))
p <- ggplot(.df, aes(time, surv, group = strata)) +
    geom_step(aes(linetype = strata), size = 1) +
    theme(legend.position = c(.65, .5), legend.key = element_rect(colour = NA)) +
    labs(linetype = ystrataname) + 
    scale_y_continuous(ylabs, limits = c(0, 1)) +
    labs(linetype = ystrataname) 
 
## Plotting the graphs
p
   
}
```



```{r plot3, echo = FALSE, fig.width = 5, fig.height = 5, fig.align = 'center', warning = FALSE}
D <- rbinom(250, 1, .3)
T <- rnorm(250, mean = D, sd = .75)
  
c <- sort(T)
TPF <- sapply(c, function(x) mean(T[D == 1] > x))
FPF <- sapply(c, function(x) mean(T[D == 0] > x))

badroc <- data.frame(cbind(c, TPF, FPF))

plot(TPF ~ FPF, badroc, lwd = 2, xlab = "1 - specificity", ylab = "sensitivity", type = 's', xaxp = c(0, 1, 2), yaxp = c(0, 1, 2))
legend("bottomright", lwd = 2, lty = 1, legend = "Great new marker\n AUC = .79; p < 0.001")
```

## Evaluating Biomarkers: Practice

```{r plot2, echo = FALSE, fig.width = 6, fig.align = 'center', warning = FALSE}
df <- data.frame(Time = c(rexp(50, .1), rexp(50, 2), rexp(50, 4), rexp(50, 3)), 
                 Biomarker = c(rep("Below median", 100), rep("Above median", 100)), 
                 Treatment = rep(c(rep("Experimental", 50), rep("Placebo", 50)), 2))


df <- ddply(df, ~ Biomarker, function(d){
  
  data.frame(Time = d$Time, Survival = (nrow(d):1/nrow(d))[rank(d$Time, ties.method = "first")])
  
})

ggplot(df, aes(x = Time, y = Survival, color = Biomarker)) + scale_x_continuous(limits = c(0, 10)) +
  geom_step() + theme_igray() 

```
* p < 0.001


## Evaluating Biomarkers: Practice

```{r plot1, echo = FALSE, fig.align='center', warning=FALSE}

df <- data.frame(Time = c(rexp(50, .1), rexp(50, 2), rexp(50, 4), rexp(50, 3)), 
                 Biomarker = c(rep("Below median", 100), rep("Above median", 100)), 
                 Treatment = rep(c(rep("Experimental", 50), rep("Placebo", 50)), 2))

df <- ddply(df, ~ Biomarker + Treatment, function(d){
  
  data.frame(Time = d$Time, Survival = (nrow(d):1/nrow(d))[rank(d$Time, ties.method = "first")])
  
})

ggplot(df, aes(x = Time, y = Survival, color = Treatment)) + scale_x_continuous(limits = c(0, 10)) +
  geom_step() + theme_igray() + scale_color_colorblind() + facet_wrap(~Biomarker)
```
* p < 0.001



## Evaluating Biomarkers: Theory


Search Term | Results
-----------|-------------
Clinical Trial(s) | 926
Survival | 714
Longitudinal | 634
**(Bio)marker(s)** | **228**
Gene/Genomic(s) | 180


Search results from pubmed restricted to journals: _Statistics in Medicine, Biometrics, Biostatistics, Biometrical Journal_


## About me and the BRB

Responsibilities:

  - Provide statistical leadership to programs in the CDP
  - Conduct research in Biostatistics
      + Not quite mathematics
      + Not quite epidemiology
      + Methods are intended to be used/understood by statisticians and non-statisticians

**Much of the published statistical research will never be used in practice.**

## Why?

  - Lack of awareness
  - Lack of useful tools to implement the research
  
**Greater impact through software and education**
 
Some of my efforts on this front can be found here:

http://sachsmc.github.io
 

## Goals

 - Review statistical methods for evaluating biomarkers
 - Review practical needs of scientists evaluating biomarkers
 - Examine how the two intersect
 - Provide resources for using non-standard methods

## Uses of biomarkers/imaging tests in oncology

  - Screening (mammography, HPV test)
  - Diagnosis (HER-2, PSA)
  - Prognosis, predicting future recurrence (Oncotype, Mammaprint)
  - Treatment selection (ER, ALK, KRAS)
  
[Source (cancer.gov)](http://www.cancer.gov/cancertopics/factsheet/detection/tumor-markers)  
 
# The Basics

## Biomarker statistics 101

Binary biomarker and Binary outcome

<style>
td { border: 1px #aaa solid;
     padding: .2em; 
     }
</style>

<table align="center" style="text-align:center; border:1px; background:transparent; font-size: medium;">
<tr>
<td colspan="2" style="border:none;"></td>
<td colspan="2" style="background:#eeeebb;"><b>Condition<br />
(as determined by "Gold standard")</b></td>
</tr>
<tr>
<td style="border:none;"></td>
<td style="background:#ddddcc;">Total population</td>
<td style="background:#ffffcc;">Condition positive</td>
<td style="background:#ddddaa;">Condition negative</td>
<td style="background:#ddddcc;">Prevalence =<br />
Σ&#160;Condition positive
<div style="border-top:1px solid;">Σ&#160;Total population</div>
</td>
</tr>
<tr>
<td rowspan="2" style="background:#bbeeee;"><b>Test<br />
outcome</b></td>
<td style="background:#ccffff;">Test<br />
outcome<br />
positive</td>
<td style="background:#ccffcc;"><span style="color:#006600;"><b>True positive</b></span></td>
<td style="background:#eedddd;"><span style="color:#cc0000;"><b>False positive</b></span><br />
</td>
<td style="background:#ccffff;"> Positive predictive value (PPV) =<br />
Σ True positive
<div style="border-top:1px solid;">Σ&#160;Test&#160;outcome&#160;positive</div>
</td>
</tr>
<tr>
<td style="background:#aadddd;">Test<br />
outcome<br />
negative</td>
<td style="background:#eedddd;"><span style="color:#cc0000;"><b>False negative</b></span><br />
</td>
<td style="background:#bbeebb;"><span style="color:#006600;"><b>True negative</b></span></td>
<td style="background:#aadddd;">Negative predictive value (NPV) =<br />
Σ True negative
<div style="border-top:1px solid;">Σ&#160;Test&#160;outcome&#160;negative</div>
</td>
</tr>
<tr>
<td style="border:none;"></td>
<td style="border:none;"></td>
<td style="background:#ffffcc;">True positive fraction (TPF, Sensitivity) =<br />
Σ True positive
<div style="border-top:1px solid;">Σ&#160;Condition&#160;positive</div>
</td>
<td style="background:#ddddaa;">False positive fraction (FPF) =<br />
Σ False positive
<div style="border-top:1px solid;">Σ&#160;Condition&#160;negative</div>
</td>
<td style="background:#ddddcc;">Accuracy (ACC) =<br />
Σ&#160;True positive + Σ True negative
<div style="border-top:1px solid;">Σ&#160;Total population</div>
</td>
</tr>
<tr>
<td style="border:none;"></td>
<td style="border:none;"></td>
<td style="background:#ffffcc;">False negative fraction (FNF) =<br />
Σ False negative
<div style="border-top:1px solid;">Σ&#160;Condition&#160;positive</div>
</td>
<td style="background:#ddddaa;">True negative fraction</a> (TNF, Specificity) =<br />
Σ True negative
<div style="border-top:1px solid;">Σ&#160;Condition&#160;negative</div>
</td>
</tr>
<tr>
<td style="border:none;"></td>
</tr>
</table>

## Probabilities and Proportions

**Quality of test-based classifications:**

$$ TPF = Pr\{\mbox{Marker positive} | \mbox{Has condition} \} = Pr\{M = 1 | D = 1\} $$

$$ FPF = Pr\{M = 1 | D = 0 \} $$ 

Ideal: TPF = 1, FPF = 0; Bad: TPF = FPF

**Predictions**

$$ PPV = Pr\{D = 1 | M = 1\} $$

$$ NPV = Pr\{D = 0 | M = 0\} $$
 
Ideal: PPV = NPV = 1; Bad: PPV = $Pr\{D = 1\}$; NPV = $Pr\{D = 0\}$
 
## Aside: terminology

 - Sensitivity = <emph class="blue">True</emph> <emph class="red">positive</emph> fraction
 - Specificity = 1 - <emph class="blue">False</emph> <emph class="red">positive</emph> fraction
 - $Pr\{\color{red}{M = 1} | \color{blue}{D = 1}\}$
 - $Pr\{\color{red}{M = 1} | \color{blue}{D = 0}\}$
 
### You may also see FPR/TPR for false/true positive **rate**. 
### These are probabilities or proportions, not rates. 
 
 
## Examples

> Approximately 50% of women screened annually for 10 years in the United States will experience abnormal mammogram but have no cancer

> Mammograms miss about 20 percent of breast cancers that are present at the time of screening

[Source (cancer.gov)](http://www.cancer.gov/cancertopics/factsheet/detection/mammograms)

> - What is the population under study (general)? 
> - What does the test do? 
> - What sub-population does the statistic apply to? (Test +, test -, healthy, diseased)

## One statistic not enough

**All of the information in that table can be obtained given only 3 of the quantities: **

 - Usually, $(TPF, FPF, Pr\{D = 1\})$
 - Occassionally, $(PPV, NPV, Pr\{M = 1\})$

Why? Bayes Rule:

$$ Pr\{M = 1 | D = 1\} = \frac{Pr\{M = 1 \mbox{ AND } D = 1\}}{Pr\{D = 1\}} = \frac{Pr\{D = 1 | M = 1\} Pr\{M = 1\}}{Pr\{D = 1\}} $$
 
 - Most commonly, only a HR or OR is reported
 - HR/OR are insufficient to evaluate the accuracy of a test ([Pepe et al. 2004 AJE](http://aje.oxfordjournals.org/content/159/9/882.short))



## Extension 1

What if the test yields a continuous measurement?

```{r mamm, echo = FALSE, warning = FALSE, message = FALSE}
contex <- data.frame(Result = c(exp(rnorm(250)), rbinom(250, 1, .15)), Test = c(rep("Prostate specific antigen", 250), rep("Mammogram", 250)))
ggplot(contex, aes(x = Result)) + geom_bar(stat = "bin") + facet_wrap(~ Test, scales = "free") + theme_igray() + scale_color_colorblind()
```

## Solutions

### Goal is to assess accuracy using the aforementioned measures

> - Pick an arbitrary cutoff, say $c$. 
> - Now *test positive* is defined as $M > c$
> - Consider measures as functions of $c$, i.e.
    $$ TPF(c) = Pr\{M > c | D = 1\} $$
    $$ FPF(c) = Pr\{M > c | D = 0\} $$
    
## ROC Curve

```{r roctour, fig.width = 5, fig.height = 5, fig.align = 'center', echo = FALSE}
D <- rbinom(250, 1, .3)
M <- rnorm(250, mean = D, sd = .75)

rocme <- function(M, D){
  
c <- sort(M)
TPF <- sapply(c, function(x) mean(M[D == 1] > x))
FPF <- sapply(c, function(x) mean(M[D == 0] > x))

data.frame(cbind(c, TPF, FPF))

}

tour <- rocme(M, D)
ggplot(tour, aes(x = FPF, y = TPF)) + geom_path() + geom_abline(intercept = 0, slope = 1, lty = 2) + theme_igray() + 
  scale_x_continuous("False positive fraction at c") + scale_y_continuous("True positive fraction at c") + 
  annotate("text", x = .5, y = .5, label = "Reference line", hjust = 0, vjust = 1)
```

## Comparing ROC Curves

```{r roccomp, fig.width = 7, fig.height = 5, fig.align = 'center', echo = FALSE}
D <- rbinom(250, 1, .3)
M1 <- rnorm(250, mean = D, sd = .75)
M2 <- rnorm(250, mean = D, sd = 1.75)

tour2 <- rbind(rocme(M1, D), rocme(M2, D))
tour2$Marker <- c(rep("New great", 250), rep("Old bad", 250))

ggplot(tour2, aes(x = FPF, y = TPF, color = Marker)) + geom_path() + geom_abline(intercept = 0, slope = 1, lty = 2) + 
  theme_igray() + scale_color_colorblind() + 
  scale_x_continuous("False positive fraction at c") + scale_y_continuous("True positive fraction at c") 
```

## 

<iframe src="https://sachsmc.shinyapps.io/illustrateROC/"></iframe>


## Make your own ROC Curves

http://sachsmc.shinyapps.io/plotROC/

Send me your feedback. 

## Extension 2

  - Binary outcome, good for diagnosis, screening
  - What about prognostic/predictive biomarkers?
  - Outcome of interest is *time to event*, i.e. PFS or OS
  - Instead of $D = 1$ we have $T > t$
  
```{r surv, fig.height = 3, fig.width = 5, echo = FALSE, fig.align = 'center'}
ggplot(data.frame(T = sort(exp(rnorm(100, mean = 1, sd = .2)))), aes(x = T, y = 1 - seq_along(T)/100)) + geom_step() + 
  scale_x_continuous("Time to progression") +
  scale_y_continuous("Proportion progression free") + theme_igray()
```
  
## Solutions

> - Dichotomize time and proceed as binary
>     + possibly throwing out data (censoring)
> - Use the Kaplan-Meier estimate 
>     + Time-dependent TPF and FPF
      $$ TPF(c, t) = P(M > c | T \leq t) $$
      $$ FPF(c, t) = P(M > c | T > t) $$

## Time dependent TPF and FPF

A little algebra (Bayes rule again): 

$$ TPF(c, t) = \frac{P(T \leq t | M > c) P(M > c)}{P(T \leq t)} $$

The KM curve is an estimate of $P(T > t)$, combine that with the proportion of $M > c$ and we are done. 

## The plot
Show curves for a small subset of event times:

```{r survplot, echo = FALSE, warning = FALSE, fig.width = 5, fig.height = 4}

survT <- rexp(350, 1/5)
cens <- rbinom(350, 1, .1)

M <- -8 * sqrt(survT) + rnorm(350, sd = survT)

marker <- factor(M < median(M), labels = c("above median", "below median"))

sfit <- survfit(Surv(survT, cens) ~ marker)

sroc <- lapply(c(2, 5, 10), function(t){ 
  stroc <- survivalROC(Stime = survT, status = cens, marker = M, predict.time = t, method = "NNE", span = .25 * 350^(-.2))
  data.frame(TPF = stroc$TP, FPF = stroc$FP, time = rep(stroc$predict.time, length(stroc$FP)))
  })

multi_ggroc(sroc, label = paste("t =", c(2, 5, 10)), label.adj.y = rep(.07, 3)) + theme_igray() + geom_abline(slope = 1, intercept = 0, col = "gray90")
ggkm(sfit, timeby = 10) + scale_x_continuous(limits = c(0, 15)) + theme_igray() + theme(legend.position = c(.75, .25))


```


## Prediction/Treatment Selection
The Janes lab at the Fred Hutch has done lots of work in this area. `TreatmentSelection` R package, and others. 


```{r trtsel, include = FALSE, echo = FALSE, message = FALSE, warning = FALSE}
library(TreatmentSelection)
data(tsdata)
trtsel.Y1 <- trtsel(event = "event", 
                    trt = "trt", 
                    marker = "Y1", 
                    data = tsdata, 
                    study.design = "randomized cohort",
                    link = "logit", 
                    default.trt = "trt all")

trtsel.Y2_disc <- trtsel(event = "event", 
                         trt = "trt", 
                         marker = "Y2_disc", 
                         data = tsdata, 
                         study.design = "randomized cohort", 
                         link = "logit")

p1 <- plot.trtsel(trtsel.Y2_disc,
                   main = "Discrete version of Y2", 
                   plot.type = "risk", 
                   ci = "vertical", 
                   conf.bands = TRUE, 
                   offset = 0.01, 
                   bootstraps = 50, 
                   trt.names = c("chemo.", "no chemo."))

p2 <- plot.trtsel(trtsel.Y1, 
            main = "Y1: Oncotype-DX-like marker", 
            plot.type = "risk", 
            ci = "horizontal", 
            conf.bands = FALSE, 
            bootstraps = 50,       #more bootstraps should be run than this in practice!
            trt.names = c("chemo.", "no chemo."))
```

```{r intpult, fig.width = 5, fig.height = 4, echo = FALSE, message = FALSE, warning = FALSE}
p2$plot + theme_igray() + scale_linetype_discrete("Treatment", labels = c("Chemo", "No Chemo"))
p1$plot + theme_igray() + scale_linetype_discrete("Treatment", labels = c("Chemo", "No Chemo")) + 
  scale_shape_discrete("Treatment", labels = c("Chemo", "No Chemo"))

```
<p class="footnote">
Janes, Holly; Brown, Marshall D; Pepe, Margaret; Huang, Ying; "An Approach to Evaluating and
Comparing Biomarkers for Patient Treatment Selection" 2014. The International Journal of Biostatistics.
</p>
    
## Planning trials

https://mdbrown.shinyapps.io/PowerSAM/


# Conclusion

## The tools are out there

  - Statistical methods and tools are available to address the tough clinical questions
  - Don't limit to what you can do or know how to do
  - Talk to your friendly, local statistician to find what you/your colleagues need
  - My job is to make these more understandable and accessible
  
#### michael.sachs@nih.gov


